NMRlipids Databank: Making data-driven analyses of membrane properties accessible for all

### 2022-12-06 main.tex
Update main.tex

Cellular membranes are composed of hundreds of different types of lipid molecules which regulate membrane properties, morphology and biological functions~\cite{vanmeer08}. Membrane lipid composition can be also related to diseases, such as cancer and neurodegenerative disorders, and potential therapeutics that affect membrane compositions have been proposed~\cite{torres21}. However, membrane containing systems are often difficult to study experimentally because membranes are in disordered fluid state in biological conditions and sample preparation can be highly complicated due to rich phase behaviour and complex interactions between lipids and other biomolecules. Therefore, connections between complex lipid interactions and biological functions are still poorly understood.
%why such variety of lipids is needed and how their. 
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, yet their accuracy is often compromised by the quality of model parameters and other artefacts~\cite{antila22b,gupta22}. 
%While MD simulations of systems with few biomolecules continue having major contributions in our understanding on biology and drug design, 
On the  other hand, the accuracy of models is becoming increasingly important as researches are progressing from the simulations of 
%now searching new avenues by building models not only 
individual molecules to the models of whole cells or organelles using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. In such systems exhibiting very complex behaviour, inaccuracies are more difficult to detect and multiplication of small errors may have major effects the conclusions.

In contrast to experimental structural biology where standard protocols to share and quality evaluate the resolved structures are established in the PDB databank~\cite{montelione13}, such best practices are yet to be defined for MD simulations. Importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and solutions for sharing data are emerging for proteins in solution~\cite{meyer10,kamp10} or membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, cyclodextrins~\cite{mixcoha16}, and arbitrary biomolecules~\cite{bekker20}. However,
%The relevance of quality evaluation of simulation trajectories in databanks regarding technical details of simulations and accuracy of the underlying physical description of the system (force field) has become evident~\cite{tai04,meyer10,hospital20} and such 
automatic quality evaluation~\cite{meyer10,hospital16} and programmatic access are still rare. 
%However, straightforward quality comparisons between individual simulations or force fields within these databanks remain challenging. 
Therefore, tools for automatic quality evaluation of membrane simulations or training sets for machine learning models of membrane containing systems are not yet available.

@@

Here we present the NMRlipids databank, a community driven open access databank with programmatic access containing atomistic resolution MD simulations of lipid bilayers. The programmatic access enables various data-driven approaches, providing new tools for researchers in wide range of fields in academia and industry from cell membrane biology to lipid nanoparticle formulations and data-driven computational chemistry and machine learning. Here we demonstrate how data-driven analysis of water anisotropic diffusion in membrane systems from the NMRlipids databank can extend the scope of MD simulations to MRI imaging and pharmacokinetics, and how the NMRlipids databank can be utilized to analyse rare phenomena that are beyond the scope of standard MD simulation studies. Furthermore, we perform automatic quality evaluation of membrane simulations which guides the selection of best models for specific applications and development of simulation parameters and methodology.
%The practical relevance of the NMRlipids databank is exemplified by automatic quality evaluation and ranking of large amount of MD simulation data,

While the NMRlipids databank currently contains only bilayer systems, it can be readily expanded to contain other molecules than lipids, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The combination of overlay databank structure and open collaboration introduced here can be applied also when building databanks in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Therefore, the approach opens up possibilities to create databanks enabling data-driven and machine learning applications in new fields.

>>>>

Cellular membranes are composed of hundreds of different types of lipid molecules which regulate membrane properties, morphology and biological functions~\cite{vanmeer08}. Membrane lipid composition can be also related to diseases, such as cancer and neurodegenerative disorders, and potential therapeutics that affect membrane compositions have been proposed~\cite{torres21}. However, membrane-containing systems are often difficult to study experimentally because the membranes are in a disordered fluid state under biological conditions and sample preparation can be arduous due to the complicated phase behaviour and rich interactions between lipids and other biomolecules. For those reasons, the detailed connections between lipid interactions and biological functions taking place in or around membranes are still poorly understood.
%why such variety of lipids is needed and how their. 
Molecular dynamics (MD) simulations have been particularly useful in understanding membrane systems, yet their accuracy is often compromised by the quality of model parameters, methodological inadequacy, and other artefacts~\cite{antila22b,gupta22}. 
%While MD simulations of systems with few biomolecules continue having major contributions in our understanding on biology and drug design, 
On the  other hand, the accuracy of models is becoming increasingly important as the scale (in both size and time) of the systems studied continuously increases 
%researches are progressing from the simulations of 
%now searching new avenues by building models not only 
from the level of individual molecules to models of whole cells or organelles using interdisciplinary approaches~\cite{johnson15,thornburg22,gupta22}. Such systems exhibit complex emergent behaviour making inaccuracies more difficult to detect and accumulation of even modest errors may have major effects the conclusions.

In contrast to experimental structural biology, where standard protocols to share and quality evaluate the resolved structures are established in the PDB databank~\cite{montelione13}, the equivalent best practices are yet to be defined for MD simulations. The importance of such approaches is widely recognized~\cite{feig99,tai04,silva06,abraham19,hildebrand19,hospital20,abriata20,espigares20} and solutions for sharing data are emerging for proteins in solution~\cite{meyer10,kamp10}, proteins in membranes~\cite{newport19,espigares20,leston22}, nucleic acids~\cite{hospital16}, cyclodextrins~\cite{mixcoha16}, and arbitrary biomolecules~\cite{bekker20}. However,
%The relevance of quality evaluation of simulation trajectories in databanks regarding technical details of simulations and accuracy of the underlying physical description of the system (force field) has become evident~\cite{tai04,meyer10,hospital20} and such 
a streamlined workflow with automatic quality evaluation~\cite{meyer10,hospital16} and programmatic access is still rare. 
%However, straightforward quality comparisons between individual simulations or force fields within these databanks remain challenging. 
Therefore, tools for automatic quality evaluation of membrane simulations or training sets for machine learning models of membrane-containing systems are not yet available.

@@

Here we present the NMRlipids databank, a community-driven open access databank with programmatic access containing atomistic resolution MD simulations of lipid bilayers. The programmatic access enables various data-driven approaches, providing new tools for researchers in a wide range of fields covering academia and industry from cell membrane biology to lipid nanoparticle formulations, data-driven computational chemistry, and machine learning. Here, we demonstrate how data-driven analysis of water anisotropic diffusion in membrane systems from the NMRlipids databank can extend the scope of MD simulations to MRI imaging and pharmacokinetics as well as how the NMRlipids databank can be utilized to analyse rare phenomena that are beyond the scope of standard MD simulation investigations. Furthermore, we perform automatic quality evaluation of membrane simulations, which guides the selection of best models for specific applications and the improvement of simulation parameters and methodology.
%The practical relevance of the NMRlipids databank is exemplified by automatic quality evaluation and ranking of large amount of MD simulation data,

While the NMRlipids databank currently contains only bilayer systems, it can readily be expanded to contain other molecules than lipids, such as disordered proteins or sugars. 
%a solution for lipid bilayers based on overlay databank structure illustrated in Fig.~\ref{fig:overlay}.  
The combination of an overlay databank structure and a community-wide open collaboration used here has been applied successfully when building databanks in other fields,  
%is developed here to solve the practical challenges in generating databanks of MD simulation data enabling flexible analyses over large sets of simulation data, but it can be  potentially used for wide range of situations. 
particularly when storage of raw data requires significant resources, best practices in the field are not defined, and incentives to share data do not exist.
%overlay databank approach lowers the barrier to start without compromizing the long term stability or scalability.
Thus, the approach has proven itself in enabling data-driven and machine learning applications in new fields.
